---
title: "hba1c_analysis"
author: "Fang Wang"
date: "2025-12-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Predicting HbA1c with Linear Regression and LASSO (Clinical Case Study)

## 1. Clinical Motivation
*Accurately predicting HbA1c is clinically important for identifying individuals at risk of diabetes and metabolic disease. Because health data often include multiple correlated predictors, careful model selection is essential to balance prediction accuracy and interpretability.*

## 2. Data Overview
*The data used in this case study were obtained from a community health screening initiative conducted through primary-care clinics in the southeastern United States.*

## 3. Modeling Strategy
*In this case study, I compare a baseline multiple linear regression model with LASSO regression to address potential multicollinearity and improve predictive performance. Model selection is performed using 10-fold cross-validation, and performance is assessed using MSPE on both the training data and an external validation dataset.*

## 4. Baseline Model: Multiple Linear Regression
*As a baseline approach, a multiple linear regression model was fitted using all available predictors. Standard diagnostic plots were examined to assess key modeling assumptions.*

## 5. Regularization: LASSO with Cross-Validation
*To address potential multicollinearity and reduce model complexity, LASSO regression was applied. The tuning parameter was selected using 10-fold cross-validation, considering both the minimum cross-validation error and the one-standard-error rule.*

## 6. Model Performance
*Model performance was evaluated using mean squared prediction error (MSPE) and residual diagnostics. Across models, residual plots did not indicate major violations of linear modeling assumptions. MSPE values were used to compare predictive accuracy between the baseline and regularized models.*

## 7. External Validation
*To evaluate generalizability, I apply the models trained on the original dataset to an independent validation dataset. Out-of-sample MSPE is used as the primary metric to compare predictive performance across models.*

## 8. Discussion: Accuracy vs Parsimony
## 9. Conclusions


## Data and Setup 
```{r}
library(tidyverse)
library(glmnet)

hba1c = read.csv("data/HbA1c.csv")
hba1c_val = read.csv("data/HbA1c_val.csv")
```

## 4.1. Baseline Model: Multiple Linear Regression and Diagnostics

```{r}
fit_1a = lm(hba1c ~ age + bmi + physical_activity + diet_score + digit_ratio, data = hba1c)
summary (fit_1a)

png("figures/lm_diagnostics.png", width = 900, height = 900)
par(mfrow = c(2, 2))
plot(fit_1a)
dev.off()
```

Model diagnostics did not reveal major violations of linear regression assumptions. Residuals showed no clear nonlinearity, the scale-location plot suggested roughly constant variance, and the Q–Q plot indicated approximate normality. No single observation appeared highly influential based on leverage and Cook’s distance.

## 5.1. Regularized Modeling: LASSO with Cross-Validation

```{r}
# using cross validation to choose lambda:
lambda_seq <- 10^seq(-3, 0, by = .1)
set.seed(2025)
cv_lasso <- cv.glmnet(as.matrix(hba1c[2:6]), hba1c$hba1c, 
                       alpha = 1,
                       lambda = lambda_seq, 
                       nfolds = 10)

png("figures/lasso_cv.png", width = 800, height = 600)

plot(cv_lasso)
abline(
  v = log(cv_lasso$lambda.min),
  col = "blue",
  lwd = 2,
  lty = 2
)
abline(
  v = log(cv_lasso$lambda.1se),
  col = "red",
  lwd = 2,
  lty = 2
)

dev.off()



# fit LASSO models using lambda.min and lambda.1se:
fit_min <- glmnet(as.matrix(hba1c[2:6]), hba1c$hba1c, 
                       alpha = 1,
                       lambda = cv_lasso$lambda.min)
coef(fit_min)

fit_1se <- glmnet(as.matrix(hba1c[2:6]), hba1c$hba1c, 
                       alpha = 1,
                       lambda = cv_lasso$lambda.1se)
coef(fit_1se)
```

Using 'lambda.min', LASSO retained all predictors with shrunken coefficients. Using 'lambda.1se', LASSO selected a simpler model retaining 'age' and 'BMI' only, improving interpretability at the cost of higher prediction error.

## 6.1 Model Performance on Training Data

```{r}
# Calculate MSPE(mean squared prediction error) for the full model:
predictions_1 <- predict(fit_1a)
residuals_1 <- hba1c$hba1c - predictions_1
mspe_1 <- mean(residuals_1^2)
mspe_1

# Calculate MSPE(mean squared prediction error) for fit_min:
predictions_2 <- predict(fit_min, newx = as.matrix(hba1c[2:6]))
residuals_2 <- hba1c$hba1c - predictions_2
mspe_2 <- mean(residuals_2^2)
mspe_2

# Calculate MSPE(mean squared prediction error) for fit_1se:
predictions_3 <- predict(fit_1se, newx = as.matrix(hba1c[2:6]))
residuals_3 <- hba1c$hba1c - predictions_3
mspe_3 <- mean(residuals_3^2)
mspe_3

# plot fits vs. residuals for the full model:
par(mfrow = c(1, 3))
plot(predictions_1, residuals_1,
     main = "Full Model",
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 19)
abline(h = 0, col = "red", lty = 2)

plot(predictions_2, residuals_2,
     main = "LASSO_lambda.min",
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 19)
abline(h = 0, col = "red", lty = 2)

plot(predictions_3, residuals_3,
     main = "LASSO_lambda.1se",
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 19)
abline(h = 0, col = "red", lty = 2)
```

## 7.1 External Validation on Independent Dataset

```{r}
pred_full = predict(fit_1a, newdata = hba1c_val)
pred_lasso_min = predict(fit_min, newx = as.matrix(hba1c_val[2:6]))
pred_lasso_1se = predict(fit_1se, newx = as.matrix(hba1c_val[2:6]))

mspe_full = mean((hba1c_val$hba1c - pred_full)^2)
mspe_full
mspe_lasso_min = mean((hba1c_val$hba1c - pred_lasso_min)^2)
mspe_lasso_min
mspe_lasso_1se = mean((hba1c_val$hba1c - pred_lasso_1se)^2)
mspe_lasso_1se
```

Across both the training set and external validation set, 'LASSO (lambda.min)' achieved the lowest MSPE, indicating the best predictive performance in this case study.The 'lambda.1se' model was more parsimonious but showed a noticeable increase in MSPE.


## **Table 1.** Comparison of predictive performance across models.

```{r}
results <- data.frame(
  Model = c("Full linear regression", "LASSO (lambda.min)", "LASSO (lambda.1se)"),
  Train_MSPE = c(mspe_1, mspe_2, mspe_3),
  Validation_MSPE = c(mspe_full, mspe_lasso_min, mspe_lasso_1se),
  Num_Predictors = c(5, 5, 2)
)
knitr::kable(results, digits = 4)
```


## 8. Discussion: Accuracy vs Parsimony
The results illustrate a common trade-off in clinical predictive modeling. While the LASSO model using the 1SE rule offers a highly interpretable solution by retaining only age and BMI, this parsimony is associated with a noticeable loss in predictive accuracy.

In contrast, the lambda.min LASSO model demonstrates superior performance in both internal and external validation. In a clinical screening or population health context—where accurate identification of individuals at risk is the primary objective—the modest increase in model complexity is justified. Nevertheless, the 1SE model may remain useful in resource-limited or communication-focused settings where simplicity is critical.

## 9. Conclusions
This case study demonstrates how regularization can improve predictive modeling of clinical biomarkers. In this setting, the LASSO model selected using lambda.min provided the best balance between accuracy and model complexity.
This project was completed as part of graduate-level data science training and is presented here as a clinical data analysis case study.
